The Neuro Bureau ADHD-200 Preprocessed Repository

### 2016-05-22 Bellec_ADHD200_2016.tex
Finished edits to the discussion. Only very minor points left to address reviews, with the notable exception of working QC inside the text (and finalize release of the QC spreadsheet).

In 2011, the ``ADHD-200 Global Competition'' was held with the aim of engaging researchers from a variety of analytical backgrounds to identify biomarkers of attention-deficit/hyperactivity disorder (ADHD) from resting-state functional magnetic resonance imaging (rs-fMRI) and structural MRI (s-MRI) data \cite{Milham2012}. The competition made use of the ``ADHD-200 Sample'' data collection that was aggregated from eight independent sites and shared through the Intenational Neuroimaging Datasharing Initiative (INDI) \cite{Mennes2013}. The data includes rs-fMRI, structural MRI (s-MRI), and basic phenotypic information for 973 individuals: 585 typically-developing controls (TDC), 362 ADHD, and 26 whose diagnosis has not been publicly released \cite{Milham2012}. Competitors were given five and a half months to optimize a classification algorithm on training data (776 individuals) and submit their predicted clinical labels on test data for which diagnostic information was withheld. The competition data was distributed in a raw form and, before any analysis could begin, the images had to be preprocessed to make them comparable across individuals and reduce noise. These preprocessing steps present a significant hurdle for would-be competitors who do not have the specialist knowledge of neuroimaging methods, or access to high performance computing resources. Realizing this barrier to entry, the Neuro Bureau prospectively collaborated with all competitors by preprocessing the data and sharing these results.

@@

\paragraph{ICN time series and spatial maps} Time series and spatial maps were derived for 10 ICNs that have been found to be consistent across resting-state datasets and a variety of neuroimaging tasks \cite{smith2009correspondence} by applying a modified dual-regression approach \cite{Beckmann2009dualreg} to the unfiltered preprocessed data. A spatial multiple regression was first used to extract time series corresponding to each network. In a second step, each time course was independently correlated with whole-brain time series to generate subject-specific functional connectivity maps for each network. Alternatively, all time series were entered simultaneously into a multiple (temporal) regression, and the regression coefficients associated with each time series constituted the functional connectivity maps. The resulting ICN time series are distributed as tab-separated values (AFNI .1D) files and the spatial maps for both temporal regression approaches are distributed as compressed 4D NIfTI files (.nii.gz).

@@

\subsubsection{Structural processing} The NIAK implements a variant of the CIVET pipeline \cite{ad2006civet}. Each individal s-MRI scan was first corrected for intensity non-uniformities \cite{sled1998nonparametric} and the brain was extracted using a region growing algorithm \cite{park2009skull}. Individual scans were then linearly registered (9 parameters) with the T1 MNI symmetric template \cite{fonov2011unbiased}, restricted to the brain with the previous mask. Note that, by selecting a symmetric template, it is possible to study functional connectivity between homotopic regions by simply flipping the $x$ axis in stereotaxic space, e.g. \cite{VMHCpap}. The s-MRI scans were again corrected for intensity non-uniformities in stereotaxic space, this time restricted to the template brain mask. An individual brain mask was extracted a second time on this improved image \cite{park2009skull} and combined with template priors. An iterative non-linear registration was estimated between the linearly registered s-MRI and the template space, restricted to the brain mask \cite{collins1994automatic}. A final brain mask of the T1 image in native space was extracted from the template brain mask by inverting the linear and non-linear transformation. This final mask was used for registration between rs-fMRI and sMRI data (see below). \emph{Shared s-MRI outputs include:} non-uniformity corrected T1 volumes in native and stereotaxic space (after linear or non-linear transformations) at 1 mm isotropic resolution and brain masks in all spaces, in compressed NIFTI format (.nii.gz), as well as the linear and non-linear transformations from native to template space, as .xfm MINC files.

@@

In line with the original purpose of the ADHD-200 Global competition, most works derived from the ADHD-200 Preprocessed initiative proposed new machine learning techniques to predict clinical diagnosis in ADHD. The most frequently used derivatives have been Athena preprocessed fMRI volumes \citep[e.g.][]{Liang2012,Carmona2015} and Athena regional time series \citep[e.g.][]{Fujita2013,Dey2014}. Some works also based their main analysis on the Burner, either in isolation \citep[e.g.][]{Rangarajan2014,Rangarajan2015,Mahanand2013} or in combination with Athena functional derivatives \citep[e.g.][]{Ji2011,Eloyan2012,Olivetti2014}. Finally, a few papers utilized the Athena ICA, ReHo and fALFF maps \citep[e.g.][]{Sato2012a} and the NIAK high resolution (either ROI1000 or ROI3000) regional time series in their main analysis \citep[e.g.][]{Ahn2015,Yang2015}. Some papers also made a more marginal use of the ADHD200 preprocessed resources than the one listed previously. First, several works only re-used the CC200 or CC400 functional brain parcellations \citep[][e.g.]{Dey2012,Dai2012}, and not the ADH200 sample itself. 

@@

The ADHD-200 sample provides ample opportunity for analyses besides the identification of rs-fMRI based biomarkers of ADHD \cite{craddock2015}. With 585 TDC participants between the ages of 7 and 26, and the inclusion of intelligence measures, the ADHD-200 Preprocessed is a valuable resource for mapping developmental trajectories \cite{dosenbach2010, zuo2010, dimartino2014} and other sources of inter-individual variation \cite{kelly2012}. Perhaps most exciting are new methods that cluster individuals based on connectivity profiles \cite{miranda2014connectotyping, finn2015fingerprinting}, which are providing new hope for using neuroimaging data to parse the heterogeneity within mental health disorders \cite{castellanos2013}. One of the outstanding needs for neuroimaging, and connectomics in particular, is the development and validation of new analytical tools and processing strategies \cite{craddock2013,varoquaux2013,craddock2015}. In the service of this aim, the ADHD-200 Preprocessed repository has the necessary components to become a benchmark dataset for evaluating new tools as they are proposed.

@@

Beyond personal ideology, or the availability of particular features, there is no clear reason to prefer one pipeline over the other. Hopefully, the presence of multiple pipelines will enable researchers to compare the robustness of their tools or analysis results to different processing choices. It isn't clear which of the parcellations provide the optimal time series, although there is growing consensus that functionally-defined brain regions represent the brain's intrinsic connectivity better than structurally-defined parcellations \cite{craddock2012whole, Blumensath2013, thirion2014}. Similarly, it isn't clear which scale of parcellation is optimal. While smaller parcels tend to be better at representing connectivity present at the voxel-scale \cite{bellec2006identification, craddock2012whole, thirion2014}, fewer parcels may be more manageable or adapted for specific algorithms \cite[e.g.]{michel2012supervised,bellec2015impact}.

@@

The ADHD-200 preprocessed has helped to expand the boundaries of the traditional neuroimaging community, with several publications in core engineering and statistics journals that do not routinely feature neuroimaging applications, such as Statistica Sinica \citep{Yang2015} or the Journal of the American Statistical Association \citep{Chen2015}, amongst others \citep[e.g.][]{Liang2012, Rangarajan2015, Rangarajan2014, Mahanand2013, Ji2011, Li2015, Li2013, Liu2012, Subramanian2013, Hou2015, Deshpande2015, She2014, He2013, Kong2013, Ahn2015}. In particular, the winning team of the ADHD-200 Global Competition was based at the Johns Hopkins Biostatistics Department and used ADHD-200 Preprocessed to develop their diagnostic algorithm \cite{Eloyan2012}. We also found interesting that a handful of methodological works used the ADHD-200 Preprocessed sample as one application in a series of benchmarks that could be as varied as positron emission tomography in Alzheimer's disease \citep{Yang2015}, the Yeast gene regulatory network \citep{She2014} or gene expression in brain tissues of patients with HIV-1 associated neurocognitive disorders \citep{Liu2012}. This suggests that fully processed samples distributed under easy-to-access formats with minimal licensing could dramatically accelerate the cross-polinization of disciplines.

@@

The impact of the ADHD-200 Preprocessed repository demonstrated the need for reducing computational barriers to participation in discovery neuroscience, including but not limited to machine learning competitions based on neuroimaging data. The ADHD-200 Preprocessed release has contributed and will continue to contribute to respond to that need. We still believe that a much larger-scale effort will be necessary to unlock the full potential of openly shared neuroimaging data in the service of accelerating neuroimaging research. In line with the grassroots, open science ethos of the Neurobureau, new contributors interested in sharing derivatives of ADHD-200 or other open imaging data repository can contact us via a forum\footnote{\url{http://www.nitrc.org/forum/?group_id=383}}. We are also actively seeking for contributions, notably during brainhack events\citep{Craddock2016}. An important current project is the ABIDE preprocessed initiative\footnote{\url{http://preprocessed-connectomes-project.org/abide/}} \citep{dimartino2014}. This new resource, still under development, will include over 15 different processing strategies for rs-fMRI processing, implemented across 4 different software packages, and 2 different pipelines for structural MRI processing. The release will feature an harmonized organization of processed derivatives across packages and extensive quality control. We are planning to continue to expand this line on work on other data sources in the near future. Our hope is that ADHD-200 Preprocessed and future related efforts will critically help fMRI researchers to identify optimal analytical paths for a given task.

>>>>

In 2011, the ``ADHD-200 Global Competition'' was held with the aim of engaging researchers from a variety of analytical backgrounds to identify biomarkers of attention-deficit/hyperactivity disorder (ADHD) from resting-state functional magnetic resonance imaging (rs-fMRI) and structural MRI (s-MRI) data \cite{Milham2012}. The competition made use of the ``ADHD-200 Sample'' data collection that was aggregated from eight independent sites and shared through the Intenational Neuroimaging Datasharing Initiative (INDI) \cite{Mennes2013}. The data includes rs-fMRI, structural MRI (s-MRI), and basic phenotypic information for 973 individuals: 585 typically-developing controls (TDC), 362 ADHD, and 26 whose diagnosis has not been publicly released \cite{Milham2012}. Competitors were given five and a half months to optimize a classification algorithm on training data (776 individuals) and submit their predicted clinical labels on test data for which diagnostic information was withheld. The competition data was distributed in a raw form and, before any analysis could begin, the images had to be preprocessed to make them comparable across individuals and reduce noise. These preprocessing steps present a significant hurdle for would-be competitors who do not have the specialist knowledge of neuroimaging methods, or access to high performance computing resources. Realizing this barrier to entry, the Neuro Bureau, a non-profit organization aimed at facilitating open science grassroots initiatives\footnote{See \url{http://www.neurobureau.org/mission-statement/} for the full mission statement. The Neurobureau is a non-profit organization registered in Germany}, prospectively collaborated with all competitors by preprocessing the data and sharing these results.

@@

\paragraph{ICN time series and spatial maps} Time series and spatial maps were derived for 10 group ICNs generated by \cite{smith2009correspondence}, which were found to be consistent across resting-state datasets and a variety of neuroimaging tasks. Based on these template ICNs, we applied a modified dual-regression approach \cite{Beckmann2009dualreg} to the unfiltered preprocessed data. A spatial multiple regression was first used to extract time series corresponding to each network. In a second step, each time course was independently correlated with whole-brain time series to generate subject-specific functional connectivity maps for each network. Alternatively, all time series were entered simultaneously into a multiple (temporal) regression, and the regression coefficients associated with each time series constituted the functional connectivity maps. The resulting ICN time series are distributed as tab-separated values (AFNI .1D) files and the spatial maps for both temporal regression approaches are distributed as compressed 4D NIfTI files (.nii.gz).

@@

\subsubsection{Structural processing} The NIAK implements a variant of the CIVET pipeline \cite{ad2006civet}. Each individal s-MRI scan was first corrected for intensity non-uniformities \cite{sled1998nonparametric} and the brain was extracted using a region growing algorithm \cite{park2009skull}. Individual scans were then linearly registered (9 parameters) with the T1 MNI symmetric template \cite{fonov2011unbiased}, restricted to the brain with the previous mask. Note that, by selecting a symmetric template, it is possible to study functional connectivity between homotopic regions by simply flipping the $x$ axis in stereotaxic space, e.g. \cite{Zuo2010-ec}. The s-MRI scans were again corrected for intensity non-uniformities in stereotaxic space, this time restricted to the template brain mask. An individual brain mask was extracted a second time on this improved image \cite{park2009skull} and combined with template priors. An iterative non-linear registration was estimated between the linearly registered s-MRI and the template space, restricted to the brain mask \cite{collins1994automatic}. A final brain mask of the T1 image in native space was extracted from the template brain mask by inverting the linear and non-linear transformation. This final mask was used for registration between rs-fMRI and sMRI data (see below). \emph{Shared s-MRI outputs include:} non-uniformity corrected T1 volumes in native and stereotaxic space (after linear or non-linear transformations) at 1 mm isotropic resolution and brain masks in all spaces, in compressed NIFTI format (.nii.gz), as well as the linear and non-linear transformations from native to template space, as .xfm MINC files.

@@

In line with the original purpose of the ADHD-200 Global competition, most works derived from the ADHD-200 Preprocessed initiative proposed new machine learning techniques and their main application on real data was to predict clinical diagnosis from the ADHD-200 Preprocessed sample. The most frequently used derivatives have been Athena preprocessed fMRI volumes \citep[e.g.][]{Liang2012,Carmona2015} and Athena regional time series \citep[e.g.][]{Fujita2013,Dey2014}. Some works also based their main analysis on the Burner, either in isolation \citep[e.g.][]{Rangarajan2014,Rangarajan2015,Mahanand2013} or in combination with the Athena functional derivatives \citep[e.g.][]{Ji2011,Eloyan2012,Olivetti2014}. A few papers utilized the Athena ICA, ReHo and fALFF maps \citep[e.g.][]{Sato2012a} and the NIAK high resolution (either ROI1000 or ROI3000) regional time series in their main analysis \citep[e.g.][]{Ahn2015,Yang2015}. Some publications also made a more marginal use of the ADHD-200 Preprocessed repository, by re-using the CC200 or CC400 functional brain parcellations \citep[e.g.][]{Dey2012,Dai2012}, and not the sample itself. 

@@

To the best of our knowledge, no published works have tried to directly compare the results generated by the Athena and NIAK pipeline. The two pipelines conceptually implement very similar steps, with key operations including the non-linear volumetric registration of individual structural MRIs in two different variants of the MNI space, and the registration of invidual BOLD and T1 images. The most striking difference lied in the reduction of structured noise: the Athena implemented a regression of a battery of nuisance variables (motion parameters, WM and GM signals), while NIAK implemented an automated labeling of structured noise in a spatial ICA \citep{perlbarg2007corsica}. The Athena strategy is fairly standard in the rs-fMRI community, while ICA-based noise attenuation techniques are less common. The other clear difference lies in the derivatives offered by the two pipelines: Athena featured a more extensive battery of maps and regional time series, while the NIAK included high-resolution (several thousands) brain parcels and associated time series. Beyond personal ideology, or the availability of particular features, there is no clear reason to prefer one pipeline over the other.  It has recently become clear that even subtle variations of processing environments can impact the end-results of processing pipelines \citep{Glatard2015-we}, even when the same processing is replicated. We thus believe it is important to check how variable conclusions drawn from MRI data are, depending of the specific software implementation used for processing. The NIAK and Athena pipeline will enable researchers to compare the robustness of their tools or analysis results to different processing choices and derivatives. 
The ADHD-200 Preprocessed sample also provides ample opportunity for analyses besides the identification of rs-fMRI based biomarkers of ADHD \cite{craddock2015}. With 585 TDC participants between the ages of 7 and 26, and the inclusion of intelligence measures, the ADHD-200 Preprocessed is a valuable resource for mapping developmental trajectories \cite{dosenbach2010, zuo2010, dimartino2014} and other sources of inter-individual variation \cite{kelly2012}. Perhaps most exciting are new methods that cluster individuals based on connectivity profiles \cite{miranda2014connectotyping, finn2015fingerprinting}, which are providing new hope for using neuroimaging data to parse the heterogeneity within mental health disorders \cite{castellanos2013}. One of the outstanding needs for neuroimaging, and connectomics in particular, is the development and validation of new analytical tools and processing strategies \cite{craddock2013,varoquaux2013,craddock2015}. In the service of this aim, the ADHD-200 Preprocessed repository has the necessary components to become a benchmark dataset for evaluating new tools as they are proposed.

@@

@@

The ADHD-200 preprocessed has helped to expand the boundaries of the traditional neuroimaging community, with several publications in core engineering and statistics journals that do not routinely feature neuroimaging applications, such as Statistica Sinica \citep{Yang2015} or the Journal of the American Statistical Association \citep{Chen2015}, amongst others \citep[e.g.][]{Liang2012, Rangarajan2015, Rangarajan2014, Mahanand2013, Ji2011, Li2015, Li2013, Liu2012, Subramanian2013, Hou2015, Deshpande2015, She2014, He2013, Kong2013, Ahn2015}. In particular, the winning team of the ADHD-200 Global Competition was based at the Johns Hopkins Biostatistics Department and used ADHD-200 Preprocessed to develop their diagnostic algorithm \cite{Eloyan2012}. We also found interesting that a handful of methodological works used the ADHD-200 Preprocessed sample as one application in a series of benchmarks that could be as varied as positron emission tomography in Alzheimer's disease \citep{Yang2015}, the Yeast gene regulatory network \citep{She2014} or gene expression in brain tissues of patients with HIV-1 associated neurocognitive disorders \citep{Liu2012}. This suggests that fully processed, easy-to-access samples could dramatically accelerate the cross-polinization of disciplines.

@@

The impact of the ADHD-200 Preprocessed repository demonstrated the need for reducing computational barriers to participation in discovery neuroscience, including but not limited to machine learning competitions based on neuroimaging data. The ADHD-200 Preprocessed release has contributed and will continue to contribute to respond to that need. We still believe that a much larger-scale effort will be necessary to unlock the full potential of openly shared neuroimaging data in the service of accelerating neuroimaging research. In line with the grassroots, open science ethos of the Neurobureau, new contributors interested in sharing derivatives of ADHD-200 or other open imaging data repository can contact us via a forum\footnote{\url{http://www.nitrc.org/forum/?group_id=383}}. We are also actively seeking for contributions, notably during brainhack events\citep{Cameron_Craddock2016-du}. An important current project is the ABIDE preprocessed initiative\footnote{\url{http://preprocessed-connectomes-project.org/abide/}} \citep{dimartino2014}. This new resource, still under development, will include over 15 different processing strategies for rs-fMRI processing, implemented across 4 different software packages, and 2 different pipelines for structural MRI processing. The release will feature an harmonized organization of processed derivatives across packages and extensive quality control. We are planning to continue to expand this line on work on other data sources in the near future. Our hope is that ADHD-200 Preprocessed and future related efforts will critically help fMRI researchers to identify optimal analytical paths for a given task.
